Demo arXiv template

### 2023-02-27 paper/paper.qmd
change color palette - yellow in the middle for diverge; more edit in HDI example

@@

A common task for drought researchers is to compute indexes at different parameter combinations. This can be used to identify the spatial and temporal extent of drought events, recommend the best parameter choice, or compare the effectiveness of indexes for monitoring drought. The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia. The purpose of the example is to demonstrate the interfaces the PACKAGE package built to allow easy computing at different parameter combinations.

@@

The function `compute_indexes()` can be used to collectively compute multiple indexes. The PACKAGE offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function. For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise). Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs. The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)). The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.

@@

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 12
#| fig-height: 5
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )), 
    aes(x = long, y = lat, color = .index)) +
  scale_color_distiller(palette = "BrBG", direction = 1, name = "Index", ) +
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

@@

```{r echo = TRUE, message=TRUE}
dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
(res <- dt %>%
  var_trans(gni_pc = log(gni_pc, base = 10)) %>% 
  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
            min = c(20, 0, 0, 2), max = c(85, 18, 15, log10(75000))) %>% 
  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
```

The output is an `indri` object with a summary of operations followed by the data output. The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`).

Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions. In the context of HDI, this can be:

  - what if arithmetic mean, or different weights, is used to reduce the three dimensions into HDI? 
  - what if an alternative maximum or minimum value is used in the rescaling step?

These questions can be answered using the function `swap_exprs()` and `swap_values()`. Both functions first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested. Then, `.expr` and `.values` will accept a list of expressions or a vector of values for testing. Lastly, the initial `indri` (raw data) needs to be provided.

*What if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI?*

A common method to reduce the data dimension is through linear combination. With different weighting schemes, the combination can be compared to study the variable importance of each dimension or to reveal the underlying structures of countries. The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively):

@@

Examples have been given in the drought indexes and human development index to demonstrate computing of indexes with different parameters combinations and how alternative index design can provide insights to understand distinctive country characteristics that could sometimes be overlooked. The accompanied package, PACKAGE, is not meant to provide comprehensive implementation for all indexes across all domains. Instead, it demonstrates implementing individual pipeline steps that are versatile to multiple indexes and composing new indexes from existing steps. Domain experts are welcomed to adopt the pipeline approach to develop specialised packages for specific-domains indexes.

>>>>

<!-- # 1. unpack or the argumebt s  -->
<!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->
<!-- # 3. evalute  -->
<!-- # 4. update the data object  -->

@@

A common task for drought researchers is to compute indexes at different parameter combinations. This can be used to identify the spatial and temporal extent of drought events, recommend the best parameter choice, or compare the effectiveness of indexes for monitoring drought. The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia. The purpose of the example is to demonstrate the interfaces the tidyindex package built to allow easy computing at different parameter combinations.

@@

The function `compute_indexes()` can be used to collectively compute multiple indexes. The tidyindex offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function. For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise). Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs. The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)). The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.

@@

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 12
#| fig-height: 5
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )), 
    aes(x = long, y = lat, color = .index)) +
  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

@@

```{r echo = TRUE, message=TRUE}
dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
res <- dt %>%
  var_trans(gni_pc = log(gni_pc, base = 10)) %>% 
  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
            min = c(20, 0, 0, 2), max = c(85, 18, 15, log10(75000))) %>% 
  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
  dim_red(index = (life_exp * sch * gni_pc)^(1/3))
```

<!-- The output is an `indri` object with a summary of operations followed by the data output. The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`). -->

Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions. In the context of HDI, this can be what if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI? Apart from the geometrical mean, linear combination is a common method to reduce the data dimension. With different weighting schemes, the combination can be compared to study the variable importance of each dimension or to reveal the underlying structures of countries. The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively) using the `swap_exprs()` function. The function first locates the operation to be tested (`.var = index`), then accepts a list of expressions for testing `.expr`, before requesting the raw data:

@@

Examples have been given in the drought indexes and human development index to demonstrate computing of indexes with different parameters combinations and how alternative index design can provide insights to understand distinctive country characteristics that could sometimes be overlooked. The accompanied package, tidyindex, is not meant to provide comprehensive implementation for all indexes across all domains. Instead, it demonstrates implementing individual pipeline steps that are versatile to multiple indexes and composing new indexes from existing steps. Domain experts are welcomed to adopt the pipeline approach to develop specialised packages for specific-domains indexes.

### 2023-02-27 paper/paper.tex
change color palette - yellow in the middle for diverge; more edit in HDI example

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

The function \texttt{compute\_indexes()} can be used to collectively
compute multiple indexes. The PACKAGE offers wrapper functions, with the
prefix idx\_, that simplify the calculation of commonly used indexes by
combining a set of pipeline steps into a single function. For example,
the function \texttt{idx\_spei()} includes the five steps previously
described in Section~\ref{sec-toy-example} (variable transformation,
dimension reduction, temporal aggregation, distribution fit, and
normalise). Each \texttt{idx\_xxx()} function specifies the relevant
parameters relevant to the index: the thornthwaite method is used to
calculate PET in SPEI, with the average temperature (\texttt{tavg}) and
latitude (\texttt{lat}) used as inputs. The SPEI is computed at four
time scales (6, 12, 24, and 36 months) and fitted with two distributions
(Log-logistic and General Extreme Value (GEV)). The SPI is also computed
at the same four time scales and uses the default gamma distribution to
fit the aggregated series.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
\end{verbatim}

Data: 
# A tibble: 191 x 8
      id country                life_exp exp_sch avg_sch gni_pc   sch index
   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
 2     2 Norway                    0.973   1       0.867  0.978 0.933 0.961
 3     3 Iceland                   0.964   1       0.918  0.955 0.959 0.959
 4     4 Hong Kong, China (SAR)    1       0.960   0.815  0.973 0.887 0.952
 5     5 Australia                 0.993   1       0.848  0.936 0.924 0.951
 6     6 Denmark                   0.944   1       0.864  0.967 0.932 0.948
 7     7 Sweden                    0.969   1       0.841  0.952 0.920 0.947
 8     8 Ireland                   0.954   1       0.772  1     0.886 0.945
 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
10    10 Netherlands               0.949   1       0.839  0.956 0.919 0.941
# ... with 181 more rows
\end{verbatim}

The output is an \texttt{indri} object with a summary of operations
followed by the data output. The data output contains the ID, country
name, rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
\texttt{avg\_sch}, and \texttt{gni\_pc}), the aggregated education
dimension (\texttt{sch}), and the final index calculated
(\texttt{index}).

Index analysts can be interested in studying the property of the index
and the countries through the ``what-if'' questions. In the context of
HDI, this can be:

\begin{itemize}
\tightlist
\item
  what if arithmetic mean, or different weights, is used to reduce the
  three dimensions into HDI?
\item
  what if an alternative maximum or minimum value is used in the
  rescaling step?
\end{itemize}

These questions can be answered using the function
\texttt{swap\_exprs()} and \texttt{swap\_values()}. Both functions first
use a set of arguments (\texttt{.var}, \texttt{.module}, \texttt{.step},
and \texttt{.res}) to locate the operation to be tested. Then,
\texttt{.expr} and \texttt{.values} will accept a list of expressions or
a vector of values for testing. Lastly, the initial \texttt{indri} (raw
data) needs to be provided.

\emph{What if an alternative dimension reduction expression is used to
reduce the three dimensions into the HDI?}

A common method to reduce the data dimension is through linear
combination. With different weighting schemes, the combination can be
compared to study the variable importance of each dimension or to reveal
the underlying structures of countries. The following code tests five
expressions (arithmetic mean, principal component analysis weight, and
three weighting schemes emphasizing life expectancy, education, and GNI
per capita respectively):

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

The function \texttt{compute\_indexes()} can be used to collectively
compute multiple indexes. The tidyindex offers wrapper functions, with
the prefix idx\_, that simplify the calculation of commonly used indexes
by combining a set of pipeline steps into a single function. For
example, the function \texttt{idx\_spei()} includes the five steps
previously described in Section~\ref{sec-toy-example} (variable
transformation, dimension reduction, temporal aggregation, distribution
fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
relevant parameters relevant to the index: the thornthwaite method is
used to calculate PET in SPEI, with the average temperature
(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
computed at four time scales (6, 12, 24, and 36 months) and fitted with
two distributions (Log-logistic and General Extreme Value (GEV)). The
SPI is also computed at the same four time scales and uses the default
gamma distribution to fit the aggregated series.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
\NormalTok{res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Index analysts can be interested in studying the property of the index
and the countries through the ``what-if'' questions. In the context of
HDI, this can be what if an alternative dimension reduction expression
is used to reduce the three dimensions into the HDI? Apart from the
geometrical mean, linear combination is a common method to reduce the
data dimension. With different weighting schemes, the combination can be
compared to study the variable importance of each dimension or to reveal
the underlying structures of countries. The following code tests five
expressions (arithmetic mean, principal component analysis weight, and
three weighting schemes emphasizing life expectancy, education, and GNI
per capita respectively) using the \texttt{swap\_exprs()} function. The
function first locates the operation to be tested
(\texttt{.var\ =\ index}), then accepts a list of expressions for
testing \texttt{.expr}, before requesting the raw data:
